---
title: "Using data.table, gt_summary, and table1 to create tables of summary statistics"
author: "George Savva"
date: "06/01/2021"
output: 
  html_document:
    theme: cosmo
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Making tables of descriptive statistics (typically the 'table 1' in your paper) is an important aspects of your data analysis and reporting.  

Unfortunately R does not provide a natural way to produce tables of descriptive statistics.  We can use base functions `aggregate()` and `table()` to produce means over groups or contingency tables, and this may be enough for you to buid your own tables, but these are not very flexible or attractive, and leave you a lot of work to do in collating the information that you need.

Many (many!) packages have been developed by others to fill this gap, in this short vignette I will look at three options for making tables of summary statistics in R.

See here for a discussion of nine different packages to create descriptive tables: (https://thatdatatho.com/2018/08/20/easily-create-descriptive-summary-statistic-tables-r-studio/)

We will use the tree data that you have already seen in the course, and will produce a neat table of the mean and standard deviation of tree height, and the proportion of healthy trees, stratified by species.

First we'll load the data:

```{r warning=FALSE}
library(readxl)
TreeData <- read_excel(path="introstat.xlsx", sheet="P1-TreeSpeciesData", .name_repair = "universal")
```


# Using the data.table package

data.table is an R package designed to make working with data frames easier.

Its central function is an extension of the standard subsetting function `[`.  This takes three aguments (instead of the usual two), that correspond to 

1. The rows we want to use
2. The calculations to perform
3. How to group the rows

Making a table of descriptives using this function is easiest illustrated using a simple example.   

First we need to load the package, then make our `data.frame` into a `data.table` with the `setDT` function:

```{r warning=FALSE}
library(data.table)
TreeData <- read_excel(path="introstat.xlsx", sheet="P1-TreeSpeciesData", .name_repair = "universal")
TreeDataDT <- setDT(TreeData)
class(TreeDataDT)
```


First we need to convert our data frame into a data.table object:

```{r }
TreeDataDT <- setDT(TreeData)
```


To get the mean tree height over species, including all rows in our dataset we would run:

```{r }
TreeDataDT[,                # nothing here means select all rows
           mean(height),    # calculate the mean of 'height'
           species.name     # stratify by species.name
           ]
```

We can add a second column for standard deviation, by using a `list` for the second argument:

```{r }
class(TreeDataDT)
TreeDataDT[,                # select all rows
           list( mean(height), 
                 sd(height)),    # calculate the mean and sd of 'height'
           species.name     # stratify by species.name
           ]
```

Now we can give each column a nice name:

```{r }
TreeDataDT[,                # select all rows
           list( `Mean Height`=mean(height), 
                 `SD` = sd(height)),    # calculate the mean and sd of 'height'
           list(`Species` = species.name)     # stratify by species.name
           ]
```

And finally add the proportion of healthy trees:

```{r }
TreeDataDT[,
           list( `Mean Height`=mean(height), 
                 `SD` = sd(height),
                 `% healthy` = 100*mean(health=="Y")),
           list(`Species` = species.name)
           ]
```

Now we'd like to round all the numeric values to 1dp:

```{r }
TreeDataDT[,                # select all rows
           list( `Mean Height`=round(mean(height),1), 
                 `SD` = round(sd(height),1),
                 `% healthy` = round(100*mean(health=="Y"),1)
               ),    # calculate the mean and sd of 'height'
           list(`Species` = species.name)     # stratify by species.name
           ]
```


This is nice as it results in another data frame (or data table) that we can use for plotting or further analysis, or can pass to another function to create a neat output.

Or since we can use any R function within this function we can organise our mean and sd for a more aesthetic or compact table.

```{r }
dt1 <- TreeDataDT[,list( `Mean height (sd)`= sprintf("%0.1f (%0.1f)", mean(height),sd(height)), 
                 `% healthy` = round(100*mean(health=="Y"),1)),   
            list(`Species` = species.name)]
dt1
```

We could use `kable` to make a pretty version of our table for a report:

```{r warning=FALSE}
library(kableExtra)
library(knitr)

kable_styling(kable(dt1 ))
```




# Using `table1`

Table 1 is a package specifically designed to produce tables of descriptives.  It creates pre-formatted tables in html format that can be included directly into papers.

We can set the variables we want to describe and the variable to stratify over using the formula interface as follows:

```{r warning=FALSE, message=FALSE}
library(table1)

table1(~ height + health | species.name , data=TreeData )

```

Now we need to specify that we want only the mean and sd for the height variable, and only the proportion healthy from the health variable:

[To be added]

This table is good to be directly pasted into a report.  See the table1 package ‘vignette’ to learn more about how this works and how to change the output if you need to.

https://cran.r-project.org/web/packages/table1/vignettes/table1-examples.html


# Using gtsummary

`gtsummary` is the newest package I will use, and builds on the new `gt` system for formatting tables.

The main function to produce a descriptive table is `tbl_summary()`:

```{r }
library(gtsummary)
tbl_summary(TreeData)
```


Now we can subset the data to select which variables to include, and use the `by`, `label` and `statistic` functions to specify which statistics to produce, how to stratify the data and how to label the columns:

```{r }
tbl_summary(TreeData[,c("species.name","height", "health")], 
                    by=species.name,
                    label=list(species.name ~ "Species", height  ~ "Height (m)", health ~ "Health"),
                    statistic=list(height~"{mean} ({sd})"))
```

This is an extremely rich and flexible package, see the vignette for more details:

http://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html

