---
title: "Using data.table to create tables of summary statistics"
author: "George Savva"
date: "06/01/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Making tables of descriptive statistics is one of the most important aspects of a data analysis.  Unfortunately R does not provide a natural way to produce such a table.  In its place, 

In this short vignette I will explore several options for making tables of summary statistics in R.  These are:



## Using `data.table`

```{r }
library(readxl)
library(data.table)
TreeData <- read_excel(path="old_files/day1_july_2020/introstat.xlsx", sheet="P1-TreeSpeciesData", .name_repair = "universal")
TreeDataDT <- setDT(TreeData)
```

data.table is an R package designed to make working with data frames easier.

It's central function is an extension of the standard subset function `[`.  This takes three aguments, that correspond to 
* The rows we want to use
* The calculations to perform
* How to group the rows

Making a table of descriptives using this function is easiest illustrated using a simple example.   To get the mean tree height over species, including all rows in our dataset we would run:

```{r }
class(TreeDataDT)
TreeDataDT[,                # select all rows
           mean(height),    # calculate the mean of 'height'
           species.name     # stratify by species.name
           ]
```

We can add a second column for standard deviation, by using a `list`:

```{r }
class(TreeDataDT)
TreeDataDT[,                # select all rows
           list( mean(height), 
                 sd(height)),    # calculate the mean and sd of 'height'
           species.name     # stratify by species.name
           ]
```

Now we need to give each column a nice name:
```{r }
TreeDataDT[,                # select all rows
           list( `Mean Height`=mean(height), 
                 `SD` = sd(height)),    # calculate the mean and sd of 'height'
           list(`Species` = species.name)     # stratify by species.name
           ]
```

This is nice as we 