---
title: "Using data.table to create tables of summary statistics"
author: "George Savva"
date: "06/01/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this short vignette I will explore several options for making tables of summary statistics in R.

## Using `data.table`

```{r }
library(readxl)
library(data.table)
TreeData <- read_excel(path="old_files/day1_july_2020/introstat.xlsx", sheet="P1-TreeSpeciesData", .name_repair = "universal")
TreeDataDT <- setDT(TreeData)
```

data.table is an R package designed to make working with data frames easier.

It's central function is an extension of the standard subset function `[`.  This takes three aguments, that correspond to a 

Once we have our data frame converted into a data.table, we can use this extended function to easily create summaries.

```{r }
class(TreeDataDT)
TreeDataDT[,                # select all rows
           mean(height),    # calculate the mean of 'height'
           species.name     # stratify by species.name
           ]
```

We can add 

